GPU-friendly Stroke Expansion

### 2024-06-12 paper.tex
Rewrite of introduction
The introduction is rewritten, both to address reviewer feedback and also to strenghten our statement of strong correctness.

Also redo the transition from circular arc to the sqrt-kappa error metric, making the statement stronger, and also including the monotonic curvature condition.

More is to be done, basically making things flow more smoothly in subsequent sections, but I wanted to get this part up for review, as other revision will depend on it.

Rendering of vector graphics documents requires handling both filled and stroked primitives. There is substantial literature on GPU rendering of filled paths, but many fewer published techniques for strokes. It is a more challenging problem, especially for parallel computation, because path segments cannot be processed independently of each other; the \emph{joins} between adjacent path segments depend on context. In addition, the path topology affects the rendered result, in particular whether subpaths are open or closed. In the former case, the endpoints are rendered with \emph{caps}. Joins and caps can have multiple styles, including different styles within the same document.

Stroke expansion is the process of generating an outline that, when filled, produces the \emph{stroke} of a given input path. Informally, a line of the same width as the stroke is swept along the path, normal to it, and also decorated with joins and caps. Though stroke rendering is referenced in many 2D graphics standards, it has lacked a precise formal definition until recent work, primarily \citet{Nehab2020} and \citet{Kilgard2020}.

The \citet{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work). It classifies techniques into \emph{local,} where each path segment generates closed geometry (which may be triangles or other primitives), and \emph{global,} where the overall result is a closed outline of the stroked path, but the partial result from each segment is in general open. Our technique is considered global in this scheme, yet allows independent processing of each path segment.

A number of factors contribute to an algorithm being ``GPU-friendly.'' In addition to simply being able to process the input segments in parallel, such an algorithm also avoids divergent control flow (avoiding the need for explicit subdivision at special events such as inflection points and cusps) and uses robust numerical techniques not subject to particular problems when evaluated using 32-bit floating point numbers.

We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins. The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth. As described in detail in \citet{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints. Standards for graphics formats, such as SVG[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior. While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.

A great number of rendering techniques for strokes have been proposed. Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke. Included in this category is polar stroking[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error. Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering. Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \citet{Nehab2020}. In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \citet{Nehab2020}, their evolutes when targeting strong correctness. For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \citet{Nehab2020}.

The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU. An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library. For producing curved outlines, the Tiller and Hanson[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers. The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}. A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.

@@

We are of course concerned with the flattening of more general curves (ultimately the parallel curve of a cubic Bézier), not simply circular arcs. It is tempting to sample the curvature and plug it into the formula above, but this can both dramatically undershoot the error (predicting that no subdivision is needed at an inflection point) and overshoot it (requiring infinite subdivision at a cusp; and cusps). Apparently, some kind of average curvature is needed.

We propose the following error metric to estimate the maximum distance between an arbitrary curve of length $\hat{s}$ and its chord.

@@

This formula is the same as the approximate error metric for circular arcs, except that instead of a constant curvature value, a norm-like average with an exponent of 1/2 is used (it is not considered a true norm because the triangle inequality does not hold). We chose this particular formulation because it tends to produce invertible error metrics, and because it works well in practice.

>>>>

Stroke rendering is an essential part of the standard vector graphics imaging model, and is thus widely used. The standard representation of paths is a sequence of cubic Bézier segments. While there are a number of published techniques for GPU rendering of filled paths, stroke rendering is more challenging to parallelize, largely because path segments cannot be processed independently of each other; rendering of the \emph{joins} between path segments depends on both adjoining segments, and the endpoints of paths (when not closed) are rendered with \emph{caps}. Joins and caps can have multiple styles, including different styles within the same document.

A typical vector graphics document may have thousands of stroked paths. A map or CAD drawing may have tens of thousands of paths, not unusually a million path segments. At this scale, the computational cost of stroke rendering on CPU may then result in loss of interactive rendering performance. Thus, we desire to improve performance by offloading stroke rendering to the GPU. To realize performance gains, the algorithm needs to be GPU-friendly, meaning among other things that it exploits the available parallelism, is work efficient, and avoids unnecessarily complex (and potentially divergent) control flow. In addition, it must remain numerically robust even when computed with 32 bit floating point numbers.

There are a number of strategies for rendering strokes. Among them are \emph{local} techniques that break down the stroke into individual closed primitives, distance field techniques (or similar techniques based on point inclusion). In this paper, we focus on \emph{stroke expansion,} the generation of an outline that, when filled, represents the rendered stroke. Implementing stroke rendering using stroke expansion enables a unified approach to path rendering downstream, rendering both filled and stroked paths. Among other benefits, such an approach avoids a large number of draw calls when stroked and filled paths are finely interleaved, as is often the case.

Correctness is another challenging aspect of stroke rendering. We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves of path segments, combined with caps and the outer contours of joins. Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion. Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, then addition \emph{evolute} segments need to be added, in addition to the parallel curves required by weak correctness, and also the inner contours of joins are required in general. The technique described in this paper is intended for strongly correct stroke rendering; it produces outlines that render the same shape as \citet{Nehab2020}.

Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths. Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula\citenum{Farouki1990}. Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation. A typical value is 0.25 device pixel, as that is close to the threshold of visibility. A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound. Some specifications for vector graphics, including PDF\citenum{PDF2008}, specify the error tolerance explicitly.

The subproblem of approximating parallel curves has spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU. An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library. For producing curved outlines, the Tiller and Hanson[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers. The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}. A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.

Previous work on stroke rendering does not fully solve the problem of fast and correct stroke rendering. Most existing implementations have correctness problems, in particular failing to draw the evolutes and inner joins as required for strong correctness; \citet{Nehab2020} provides a detailed survey. In addition, most implementations are not suitable for running on GPU, requiring sequential execution to produce watertight outlines. One GPU implementation is polar stroking (\cite{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).

Our central approach bounding the error tolerance is \emph{error metrics,} which predict the Fréchet distance error. The error metrics presented in this paper are both straightforward to compute, and also \emph{invertible}, which avoids excessive subdivision and also increases the amount of available parallelism to exploit. Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then error measured by sampling. Such sampling is often time consuming, involving iteration, and even then risks underestimating the error due to inadequate sampling.

@@

We are of course concerned with the flattening of more general curves (ultimately the parallel curve of a cubic Bézier), not simply circular arcs, so the curvature is not constant. An obvious approach to try would be using the maximum error. This would be conservative with respect to error tolerance, but also generate too much subdivision when the curve has a small region of high curvature. In the limit, the curve can contain a cusp of infinite curvature, which would entail infinite subdivision. Similarly, sampling the curvature at a single point can either underestimate or overestimate the error. The ideal approach would be a form of average curvature that is tractable to compute, and accurately predicts the flattening error. To that end, we propose the following error metric to estimate the maximum distance between an arbitrary curve of length $\hat{s}$ and its chord.

@@

This formula is the same as the approximate error metric for circular arcs, except that instead of a constant curvature value, a norm-like average with an exponent of 1/2 is used (it is not considered a true norm because the triangle inequality does not hold). This particular formulation has two strong advantages. First, it has been empirically validated to accurately predict the distance between chord and curve for a variety of curves. In particular, we believe that when curvature is monotonic, it is a tight bound both above and below (and later we will see that in our algorithm it is always evaluated on curve segments with such monotonic curvature). Second, this formula lends itself to invertible error metrics.
